Dependent Object Types

### 2012-09-03 doc/candidacy/report.tex
rev

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an if expression. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness, which
we illustrate with an example below. Another limitation of the SML
module system is that ``type sharing'' is limited to equality between
type names, not general type expressions. In effect, this means that
the transparency / opaqueness of a signature cannot be fully
fine-tuned. Their proposal introduces translucent signatures, which
allows any type to be partially or fully determined by a type
expression.

@@

\subsubsection*{ML-like Example and Scala Encoding}
Consider this example, in a ML-like language implementing the calculus
of the paper. Since modules are first-class, we can \code{flip} a coin
and return a module depending on the outcome:
\begin{lstlisting}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
\end{lstlisting}
But now, what should be the static type or signature of structure
\code{X}? In ML, in order to make the \code{if} type check, the two
branches must have equal types. In the calculus, the subsumption rule
is used to coerce both types to their least upper bound, represented
by the signature:
\begin{lstlisting}
signature XType = sig
  type T
  val v = T
  val f = T -> T
end
\end{lstlisting}

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allow {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with its methods and fields. {\it
  Further binding} enables the nested classes to be further
extended. Path-dependent types arise naturally to distinguish objects
coming from different families. In Tribe, path types can depend
simultaneously on both classes and objects. Additionally, paths can
use an ``out'' field to move from an object to the object which
surrounds it. This ``out'' feature enables ubiquitous access to an
object's family without the need to drag around family arguments.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and ``out'' selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
subclassing / further-binding with one another, as this would only bring
confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
``out'' faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the Tribe paper, using ``out'' in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

Our immediate goal is to prove type soundness of the DOT calculus. We
first tried to prove the calculus type-safe using the standard
theorems of preservation and progress~\cite{soundness,
  tapl}. Unfortunately, for the DOT calculus and any variants that we
devise, preservation doesn't hold. Indeed, we found many
counterexamples to preservation due to narrowing, i.e.\ a type becoming more
precise after substitution. However, the standard theorems of
preservation and progress are just one way to prove type soundness. We
are exploring a promising avenue to prove type safety using the
powerful method of step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
proving the calculus sound, we are also refining it.

>>>>

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an \code{if} expression. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness, which
we illustrate with an example below. Another limitation of the SML
module system is that ``type sharing'' is restricted to equality between
type names, not general type expressions. In effect, this means that
the transparency / opaqueness of a signature cannot be fully
fine-tuned. Their proposal introduces translucent signatures, which
allows any type to be partially or fully determined by a type
expression.

@@

\subsubsection*{ML-like Example and Scala Encoding}
Consider this example, in a ML-like language implementing the calculus
of the paper. Since modules are first-class, we can \code{flip} a coin
and return a module depending on the outcome:
\begin{lstlisting}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
\end{lstlisting}
But now, what should be the static type or signature of structure
\code{X}? In ML, in order to make the \code{if} expression typecheck, the two
branches must have equal types. In the calculus, the subsumption rule
is used to coerce both types to their least upper bound, represented
by the signature:
\begin{lstlisting}
signature XType = sig
  type T
  val v = T
  val f = T -> T
end
\end{lstlisting}

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allow {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with its methods and fields. {\it
  Further binding} enables the nested classes to be further
extended. Path-dependent types arise naturally to distinguish objects
coming from different families. In Tribe, path types can depend
simultaneously on both classes and objects. Additionally, paths can
use an \code{out} field to move from an object to the object which
surrounds it. This \code{out} feature enables ubiquitous access to an
object's family without the need to drag around family arguments.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and \code{out} selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
subclassing / further-binding with one another, as this would only bring
confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
\code{out} faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the Tribe paper, using \code{out} in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

Our immediate goal is to prove type soundness of the DOT calculus. We
first tried to prove the calculus type-safe using the standard
theorems of preservation and progress~\cite{soundness,
  tapl}. Unfortunately, for the DOT calculus and any variants that we
devised, preservation doesn't hold. Indeed, we found many
counterexamples to preservation due to narrowing, i.e.\ a type becoming more
precise after substitution. However, the standard theorems of
preservation and progress are just one way to prove type soundness. We
are exploring a promising avenue to prove type safety using the
powerful method of step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
proving the calculus sound, we are also refining it.
